Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-16 latex/main.tex
Rework the data preprocessing Methods
Closes #688

A classical tradeoff in bioinformatics is whether to store data 
in sample-major (FASTA) or variant-major (VCF) format, determining
whether the data can be accessed efficiently either by row or column
of the variant matrix. Because VCF Zarr is chunked in two dimensions,
we can now access either sample aligments or the variation
data for a given site across all samples efficiently.
See Czech et al.\cite{Czech2025} for more details and benchmarks on 
this dataset. 
Another advantage of Zarr over formats such as VCF is that is
is inherently ``appendable'', and large datasets can be updated
in real time

@@

\subsubsection*{Data preprocessing}

In this study, we used the SARS-CoV-2 sequences reassembled in a recent study \cite{Hunt2024}
using an amplicon-aware assembly method (named ``Viridian'').
The Viridian sequences contain far fewer artifacts compared to their publicly deposited counterparts on GenBank.
Consequently, phylogenetic tree inference using the Viridian data led to more parsimonious phylogenies,
as seen by a large reduction in the number of likely artifactual reversions \cite{Hunt2024}.

Prior to ARG inference, we preprocessed the remaining sequences as follows.
We treated low-quality bases (represented by Ns), gaps and ambiguous characters in the sequences as missing data.
When encoded as missing data, the bases do not inform sample matching (see ``The Li and Stephens model'').

During inference, we filtered out the samples with low-quality sequences and low-precision sampling dates.
We kept the quality check-passing sequences used to infer the Viridian UShER phylogeny \cite{Hunt2024}.
We skipped low-quality sequences that had more than 500 missing sites
(except when building the ``base'' ARG, when the maximum number of allowed missing bases was set to 10,000 instead).
We also skipped $\sim$40,000 samples which had a collection date of December 31, 2020.
The sampling frequency of this date is alarmingly high compared to other dates (at most 15,000 samples per day).
We suspect that this batch of samples might be unusually enriched for wrong collection dates,
likely due to arbitrary assignment of the originally reported collection date of ``2020'' to be ``31-12-2020''.
Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

Sites which are enriched for systematic sequencing errors or frequently experience homoplasy
can cause artifactual signals of recombination (REF).
The bioinformatics community has identified and curated a set of problematic sites
that are recommended to be excluded before phylogenetic inference (REF).
Here we assembled a bespoke set of such problematic sites
using the information in the Viridian sequences.
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites which had the highest mutation counts in the ARG (URL).
Only six of these sites overlapped with the ``Problematic Sites''
downloaded from the UCSC Genome Browser website (URL),
which did not have a high number of mutations in the preliminary ARG.
To infer the final ARG, we masked the problematic sites
in our custom set
\url{https://github.com/jeromekelleher/sc2ts-paper/blob/main/inference/viridian\_config.toml}.

>>>>

A classical tradeoff in bioinformatics is whether to store data 
in sample-major (FASTA) or variant-major (VCF) format, determining
whether the data can be accessed efficiently either by row or column
of the variant matrix. Because VCF Zarr is chunked in two dimensions,
we can now access either sample aligments or the variation
data for a given site across all samples efficiently.
See Czech et al.\cite{Czech2025} for more details and benchmarks on 
this dataset.  
Another advantage of Zarr over formats such as VCF is that is
is inherently ``appendable'', and large datasets can be updated
in real time.

@@

\subsubsection*{Sample preprocessing}
The collection date for a sequence is an important piece of information
for sc2ts. We used the Viridian
consensus date (``Date\_tree``), based on information from COG-UK, GISAID, and
ENA/SRA as our source. Samples that have missing or incomplete dates (e.g., 2020-01)
were omitted, as were
the $\sim$40,000 samples which had a collection date of December 31, 2020.
As the  sampling frequency of this date is much higher than other
(at most 15,000 samples per day), it seems likely that this 
batch is highly enriched for wrong collection dates.
Furthermore, manual checking (by cross-referencing the ENA and COG-UK Mutation Explorer)
confirmed that some of the samples were not collected in 2020.

The bioinformatics community has identified and curated a set of problematic sites
for SARS-CoV-2 that are enriched for systematic sequencing errors or 
high levels of homoplasy~\cite{DeMaio2020issues}. 
As these sites had not been updated to
take into account the systematic improvements made by the Viridian dataset,
we assembled a custom set of sites to mask. 
To do so, we first built a preliminary ARG from the samples collected up to June 30, 2021,
without any site masking.
Then, we identified the top 100 sites which had the highest mutation counts in
the ARG.
Only 11 of these sites 
(635, 8835, 11074, 11083, 15521, 16887, 21304, 21305, 21575, 21987 and 28253)
were present in the problematic sites file
(\url{https://raw.githubusercontent.com/W-L/ProblematicSites\_SARS-CoV2/master/problematic\_sites\_sarsCov2.vcf}).

Alignments read from the VCZ Zarr file were first preprocessed to mask any 
non nucleotide characters (A,C,G or T) as missing data. 
Then, we filtered 
any samples that had missing data at more than 500 sites, excluding
the 100 sites identified as problematic (this was 
set to 10,000 sites for the period up to 2020-03-01).
